### 2024-09-13 teaching/math240f24/draft.tex
BCS 2025

This problem is called the {\it boolean satisfiability problem}, often abbreviated {\mc SAT}. One
way of solving it for any given $f$ is to just compute its truth table. We already know the
downside of this approach: if $f$ has $n$ variables, then its truth table will have $2^n$ rows.
Given a few minutes, you are certainly capable of writing down a formula $f$ that has $300$
variables, call them $p_1, \ldots, p_{300}$. The truth table of $f$ will have $2^{300}$ rows,
which is more than the number of atoms in the observable universe. You can learn a lot more
about {\mc SAT} in a higher-level class on computational complexity (e.g., {\mc COMP 360/362}).
\bye

@@

\datestamp{5}{}
Propositional logic allows us to work with simple declarations, but this isn't powerful enough
to express some deeper mathematical concepts. For this purpose, we now introduce the notion of
a {\it predicate}. This is a statement involving some number of variables, each of which may take
values coming from a universe $U$, such that the statement evaluates to either true or false {\sl once
all variables are assigned values}. The statement $P(n)$ given by ``$n$ is prime''
is an example of this, where $n$
can take any value in the universe $\ZZ$. An example with two variables is the
predicate $L(x,y)$ defined by ``$x$ is less than $y$.''
(A more commonly-used notation for this predicate is ``$x<y$.'')

@@

\datestamp{6}{}
We're getting to the fun part of the course now.
Further back in these notes, we already proved a few statements about sets. This was just a taste of
what's to come, as the main focus of this course is to teach you how to prove mathematical statements.
These will all be statements that can ultimately be stated in predicate logic, so using the tools
from the previous section, you can boil them down to their logical skeleton. In this section, we will
formally define what it means to prove a statement in predicate logic.

@@

Notice how we introduced the variable $x$ in the above proof.
Because we're trying to prove a ``for all'' statement, we use
the word ``let,'' to indicate that $x$ is given to us by some higher power. In fact, we should sort
of view this higher power as possibly being malicious. A very useful way to think about writing proof is
to imagine a game in which you are trying to prove a statement in predicate logic, and a supernatural adversary
is attempting to thwart you. Let's say this predicate has four variables, so the statement is
$$\exists x \, \forall y \, \forall z\, \exists w : P(x,y,z,w).$$
The variables in the statement are introduced from left to right. Each time you see a $\exists$
symbol, it's your turn. In the example above, you get to set the variable $x$ to any element of the
given universe (keeping in mind that your eventual goal is to prove $P(x,y,z,w)$). Each time there
is a $\forall$ symbol, it's the adversary's turn, and you should be prepared for whatever he throws at you.
So in our example, the adversary may set $y$ and $z$ to anything in the universe, and he knows you picked
$x$. Lastly, you get to pick $w$. If $P(x,y,z,w)$ is true, you win, and if not, the adversary wins.
Writing a proof is equivalent to describing a winning strategy for the player against the adversary.

>>>>

This problem is called the {\it boolean satisfiability problem}, often abbreviated {\mc SAT}. One
way of solving it for any given $f$ is to just compute its truth table. We already know the
downside of this approach: if $f$ has $n$ variables, then its truth table will have $2^n$ rows.
Given a few minutes, you are certainly capable of writing down a formula $f$ that has $300$
variables, call them $p_1, \ldots, p_{300}$. The truth table of $f$ will have $2^{300}$ rows,
which is more than the number of atoms in the observable universe. You can learn a lot more
about {\mc SAT} in a higher-level class on computational complexity (e.g., {\mc COMP 360/362}).

@@

\datestamp{12}{ix}
Propositional logic allows us to work with simple declarations, but this isn't powerful enough
to express some deeper mathematical concepts. For this purpose, we now introduce the notion of
a {\it predicate}. This is a statement involving some number of variables, each of which may take
values coming from a universe $U$, such that the statement evaluates to either true or false {\sl once
all variables are assigned values}. The statement $P(n)$ given by ``$n$ is prime''
is an example of this, where $n$
can take any value in the universe $\ZZ$. An example with two variables is the
predicate $L(x,y)$ defined by ``$x$ is less than $y$.''
(A more commonly-used notation for this predicate is ``$x<y$.'')

@@

We're getting to the fun part of the course now.
Further back in these notes, we already proved a few statements about sets. This was just a taste of
what's to come, as the main focus of this course is to teach you how to prove mathematical statements.
These will all be statements that can ultimately be stated in predicate logic, so using the tools
from the previous section, you can boil them down to their logical skeleton. In this section, we will
formally define what it means to prove a statement in predicate logic.

@@

Notice how we introduced the variable $x$ in the above proof.
Because we're trying to prove a ``for all'' statement, we use
the word ``let,'' to indicate that $x$ is given to us by some higher power. In fact, we should sort
of view this higher power as possibly being malicious. A very useful way to think about writing proof is
to imagine a game in which you are trying to prove a statement in predicate logic, and a supernatural adversary
is attempting to thwart you. Let's say this predicate has four variables, so the statement is
$$\exists x \, \forall y \, \forall z\, \exists w : P(x,y,z,w).$$
The variables in the statement are introduced from left to right. Each time you see a $\exists$
symbol, it's your turn. In the example above, you get to set the variable $x$ to any element of the
given universe (keeping in mind that your eventual goal is to prove $P(x,y,z,w)$). Each time there
is a $\forall$ symbol, it's the adversary's turn, and you should be prepared for whatever he throws at you.
So in our example, the adversary may set $y$ and $z$ to anything in the universe, and he knows you picked
$x$. Lastly, you get to pick $w$. If $P(x,y,z,w)$ is true, you win, and if not, the adversary wins.
Writing a proof is equivalent to describing a winning strategy for the player against the adversary.
\bye
